---
title: "Data Manipulation Practice with dplyr"
author: "Caroline Andy"
date: "9/23/2020"
output: html_document
---

```{r setup}
library(tidyverse)
options(tibble.print_min = 3)

litters_data = read_csv("./data_import_examples/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data_import_examples/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)

```

Selecting a subset of your dataframe
```{r subset}
subset1 <- select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
subset2 <- select(litters_data, group:gd_of_birth)
```

You can also specify columns to remove.
```{r remove}
subset3 <- select(litters_data, -pups_survive)
```

Select is also really helpful to rename variables.
```{r select_to_rename}
subset4 <- select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

If you only want to rename variables and don't want to specific select or select out any variables, you can just use the rename function.
```{r rename}
subset5 <- rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

There are some handy helper functions for select; read about all of them using ?select_helpers. I use starts_with(), ends_with(), and contains() often, especially when there variables are named with suffixes or other standard patterns:
```{r helpers}
subset6 <- select(litters_data, starts_with("gd"))
```

If you want to reorder your columns, you can use everything() to list everything else not explicitly mentioned but present in the dataframe. This will reorder but not drop any columns.
```{r reorder}
subset7 <- select(litters_data, litter_number, pups_survive, everything())
```

Select will always export a dataframe, even if you are only exporting one column. Sometimes this is okay, but sometimes we want to be dealing with a vector. In these cases, we will need to use the pull function. 

```{r filter}
filter(litters_data, pups_born_alive == 3)
filter(litters_data, pups_born_alive >= 2)
#this filters out any pups that survived 4 (time units)
filter(litters_data, pups_survive != 4)
filter(litters_data, group %in% c("Con7", "Con8"))
filter(litters_data, group == "Con7" & gd_of_birth == 20)
```

Though you can use filter to drop NAs, it is preferable to use drop_na from the tidyr package.
```{r drop_na}
drop_na(litters_data) 
#drop NAs from a specific column
drop_na(litters_data, pups_born_alive)
```

Sometimes you need to select columns; sometimes you need to change them or create new ones. You can do this using mutate

```{r mutate}
mutate(litters_data,
       wt_gain = gd18_weight - gd0_weight,
       group = str_to_lower(group))
```

In comparison to the preceding, arranging is actually pretty straightforward. You can arrange the rows in your data according to the values in one or more columns:

```{r arrange}
head(arrange(litters_data, group, pups_born_alive), 10)
```
