---
title: "Tidying Practice"
author: "Caroline Andy"
date: "9/23/2020"
output: html_document
---

```{r setup}
library(tidyverse)
#limiting the number of lines printed in a  tibble
options(tibble.print_min = 5)
```

Loading sas data
```{r sas}
pulse_data = 
  haven::read_sas("./data_import_examples/public_pulse_data.sas7bdat") %>%
  janitor::clean_names()
pulse_data
```

Looking at the data, we immediately notice a problem: the BDI score is spread across four columns, which correspond to four observation times. We can fix this problem using pivot_longer:

```{r pivot_longer}
pulse_tidy_data = 
  pivot_longer(
    pulse_data, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi")
pulse_tidy_data
```

This looks much better! However, now visit is an issue. The original column names were informative but we probably don’t need to keep the bdi_score_ prefix in each case. I’ll use an additional option in pivot_longer to address this:

```{r removing_prefix}
pulse_tidy_data = 
  pivot_longer(
    pulse_data, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi")

pulse_tidy_data
```

Here we will reformat changes into one piped command as opposed to multiple intermediate steps.  
```{r tidying_format}
pulse_data = 
  haven::read_sas("./data_import_examples/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>%
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") %>%
  relocate(visit) %>%
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))) %>%
  arrange(id, visit)
pulse_data
```

Practice:
In the litters data, the variables gd0_weight and gd18_weight give the weight of the mother mouse on gestational days 0 and 18. Write a data cleaning chain that retains only litter_number and these columns; produces new variables gd and weight; and makes gd a numeric variable taking values 0 and 18 (for the last part, you might want to use recode …). Is this version “tidy”?

```{r practice}
litters_clean =
  read_csv("./data_import_examples/FAS_litters.csv") %>%
  janitor::clean_names() %>%
  select(litter_number, gd0_weight, gd18_weight) %>%
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd",
    values_to = "weight") %>%
  mutate(gd = recode(gd, "gd0_weight" = 0, "gd18_weight" = 18))
```

The code below creates a tidy dataset that could result from an analysis. This is the correct format for additional analysis or visualization, but doesn’t facilitate quick comparisons for human readers.

```{r create_tibble}
analysis_result = tibble(
  group = c("treatment", "treatment", "placebo", "placebo"),
  time = c("pre", "post", "pre", "post"),
  mean = c(4, 8, 3.5, 4)
)

analysis_result
```

I can change the table to be formatted in a more visually appealing way. The below code using pivot_wider accomplishes this. 

```{r pivot_wider}
pivot_wider(
  analysis_result, 
  names_from = "time", 
  values_from = "mean")
```

Binding rows
We’ve looked at single-table non-tidy data, but non-tidiness often stems from relevant data spread across multiple tables. In the simplest case, these tables are basically the same and can be stacked to produce a tidy dataset. That’s the setting in LotR_words.xlsx, where the word counts for different races and genders in each movie in the trilogy are spread across distinct data rectangles.

```{r reading_each_table}
#creating a new column in each data frame to denote movie choice, which was originally the variable by which entries were separated across tables. 
fellowship_ring = 
  readxl::read_excel("./data_import_examples/LotR_Words.xlsx", range = "B3:D6") %>%
  mutate(movie = "fellowship_ring")

two_towers = 
  readxl::read_excel("./data_import_examples/LotR_Words.xlsx", range = "F3:H6") %>%
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("./data_import_examples/LotR_Words.xlsx", range = "J3:L6") %>%
  mutate(movie = "return_king")
```

Once each table is ready to go, we can stack them using bind_rows and tidy the result:
```{r bind_rows}
lotr_tidy = 
  bind_rows(fellowship_ring, two_towers, return_king) %>%
  janitor::clean_names() %>%
  pivot_longer(
    female:male,
    names_to = "gender",
    values_to = "words") %>%
  mutate(race = str_to_lower(race)) %>%
  select(movie, everything())
lotr_tidy
```
  
Joining datasets
There are four main ways to join dataframes x and y:
- Inner: keeps data that appear in both x and y
- Left: keeps data that appear in x
- Right: keeps data that appear in y
- Full: keeps data that appear in either x or y

We will merge FAS_pups.csv and FAS_litters.csv which share a common variable: Litter Number. 
```{r join}
#load in FAS_pups, assign column types, clean, and recode sex variable entries to be "male" and "female" instead of ambiguous numbers 
pup_data = 
  read_csv("./data_import_examples/FAS_pups.csv", col_types = "ciiiii") %>%
  janitor::clean_names() %>%
  mutate(sex = recode(sex, `1` = "male", `2` = "female")) 
#load in FAS_litters data, assign column types, clean, separate group variable into two variables (day of treatment and treatment dose), move litter number to be first in data frame, calculate a new variable (wt_gain) and modify dose variable
litter_data = 
  read_csv("./data_import_examples/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  separate(group, into = c("dose", "day_of_tx"), sep = 3) %>%
  relocate(litter_number) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    dose = str_to_lower(dose))
#join datasets using left join
fas_data = 
  left_join(pup_data, litter_data, by = "litter_number")

fas_data
```